Bayesian integration of networks without gold standards

### 2011-09-12 src/main/latex/paper/paper.tex
added pseudocode to paper

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{color}

%% DEFINE MACROS %%
%% Math macros %%
\newcommand{\prob}{\mathbb{P}}
\newcommand{\odds}{\mathbb{O}}
\newcommand{\likeli}{\mathcal{L}}
\newcommand{\p}{\text{p}}
\newcommand{\expect}{\mathbb{E}}
%% Bio macros %%
\newcommand{\species}[1]{\textit{#1}}
%% Layout macros %%
\newcommand{\remark}[1]{\begin{center}\fcolorbox{black}{yellow}{\parbox{.8\textwidth}{#1}}\end{center}}
\newcommand{\note}[1]{{\color{red}[#1]}}
\newcommand{\REF}{{\color{red}[REF]}}

@@

An alternative paradigm is semantic data integration \citep{cheung_yeasthub:_2005,smith_obo_2007,koehler_graph-based_2006}\note{Cite Pathway commons here}. These approaches aim at representing the biological information (and as much of its meaning as possible) in a computationally accessible fashion. Rather than generalizing over all types of associations between entities to infer functional interactions, each type of association is considered separately.

@@

A full proof of the above equation is provided in the supplementary material. This recursive equation can be expressed iteratively as follows:

@@

That is, the odds of the edge to exist given all the experimental measurements is the product of the Bayes factors for these measurements with the prior odds of edge existence. These odds can be converted into the equivalent probability.

As mentioned above, the calculation assumes that experimental measurements are stochastically independent. This assumption is not valid for real data, and thus introduces a potential source of error into the methodology. Lee and colleagues address this problem by introducing dependency co-efficients for their datasets, a solution which relies heavily on guesswork \citep{lee_probabilistic_2004}. Instead we focus on making our method robust towards this kind of error.

\note{TODO: Discuss value of prior odds}

@@

To initiate the algorithm as a whole, we need to generate an initial error rate vector $\theta_{\text{init}}$. It is sufficient to sample the initial values for each $\alpha_i$ and $\beta_i$ from their prior distribution.
%$Be(1,1)$, which is equivalent to the uniform distribution $\mathcal{U}_{[0,1]}$. 
Pseudocode for the complete algorithm is given in the supplementary material.

%Finally, a score analogous to the likelihood of the current state can be computed as $\pi(\theta,G|X)$.

@@

The simulated true graph is created as a scale-free graph using a preferential attachment algorithm, since most molecular-biological graphs have been shown to be approximately scale-free \citep{jeong_lethality_2001, eisenberg_preferential_2003}. A description of the algorithm can be found in the supplementary material.
Not only does the choice of scale free background graphs more closely match the topology of real biological graphs; it also poses an additional challenge for the new algorithm, since such graphs break with the assumption of a uniform prior distribution for $G$.

@@

The next problem is systematic error, also known as experimental bias. This phenomenon in particular leads to the violation of stochastic independence between datasets. We simulate this by sampling separate false negative probabilities for each interaction and false positive probabilities for each non-interacting node-pair from Beta distributions. We then use these prepared probilities to generate false positives and false negatives in the experiments, thus introducing the same bias/systematic error. Pseudocode for this method is provided in the supplementary material. \note{TODO: Make sure it is}

>>>>

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{color} 
\usepackage{float}
\usepackage{forloop}

%% DEFINE MACROS %%
%% Math macros %%
\newcommand{\prob}{\mathbb{P}}
\newcommand{\odds}{\mathbb{O}}
\newcommand{\likeli}{\mathcal{L}}
\newcommand{\p}{\text{p}}
\newcommand{\expect}{\mathbb{E}}
%% Bio macros %%
\newcommand{\species}[1]{\textit{#1}}
%% Layout macros %%
\newcommand{\remark}[1]{\begin{center}\fcolorbox{black}{yellow}{\parbox{.8\textwidth}{#1}}\end{center}}
\newcommand{\note}[1]{{\color{red}[#1]}}
\newcommand{\REF}{{\color{red}[REF]}}
%% Pseudocode macros %%
\floatstyle{ruled}
\newfloat{pseudocode}{thp}{lop}
\floatname{pseudocode}{Pseudocode}

@@

An alternative paradigm is semantic data integration \citep{cheung_yeasthub:_2005,smith_obo_2007,koehler_graph-based_2006,cerami_pathway_2010}. These approaches aim at representing the biological information (and as much of its meaning as possible) in a computationally accessible fashion. Rather than generalizing over all types of associations between entities to infer functional interactions, each type of association is considered separately.

@@

A full proof of the above equation is provided in the supplementary methods. This recursive equation can be expressed iteratively as follows:

@@

That is, the odds of the edge to exist given all the experimental measurements is the product of the Bayes factors for these measurements with the prior odds of edge existence. (The determination the prior odds are described in the supplementary methods.) These odds can be converted into the equivalent probability using the logistic function.

As mentioned above, the calculation assumes that experimental measurements are stochastically independent. This assumption is not valid for real data, and thus introduces a potential source of error into the methodology. Lee and colleagues address this problem by introducing dependency co-efficients for their datasets \citep{lee_probabilistic_2004}, a solution which we argue relies on user guesswork. Instead we focus on making our method robust towards this kind of error.

@@

To initiate the algorithm as a whole, we need to generate an initial error rate vector $\theta_{\text{init}}$. It is sufficient to sample the initial values for each $\alpha_i$ and $\beta_i$ from their prior distribution.
%$Be(1,1)$, which is equivalent to the uniform distribution $\mathcal{U}_{[0,1]}$. 
%Pseudocode for the complete algorithm is given in the supplementary methods.

\begin{pseudocode}
  \scriptsize \input{pseudocode.tex}
  \caption{The Gibbs sampling algorithm for sampling from $\pi(\theta,G|X)$.}
\end{pseudocode}

@@

The simulated true graph is created as a scale-free graph using a preferential attachment algorithm, since most molecular-biological graphs have been shown to be approximately scale-free \citep{jeong_lethality_2001, eisenberg_preferential_2003}. A description of the algorithm can be found in the supplementary methods.
Not only does the choice of scale free background graphs more closely match the topology of real biological graphs; it also poses an additional challenge for the new algorithm, since such graphs break with the assumption of a uniform prior distribution for $G$.

@@

The next problem is systematic error, also known as experimental bias. This phenomenon in particular leads to the violation of stochastic independence between datasets. We simulate this by sampling separate false negative probabilities for each interaction and false positive probabilities for each non-interacting node-pair from Beta distributions. We then use these prepared probilities to generate false positives and false negatives in the experiments, thus introducing the same bias/systematic error. Pseudocode is provided in the supplementary methods. \note{TODO: Make sure it is}
